---
title: "Data_Visualization"
author: "William Anderson"
date: "2022-09-29"
output: html_document
---

```{r, echo = FALSE, message = FALSE}
library(tidyverse)
library(ggridges)
```


We’ll be working with NOAA weather data, which is downloaded using rnoaa::meteo_pull_monitors function in the code chunk below; similar code underlies the weather dataset used elsewhere in the course. Because this process can take some time, I’ll cache the code chunk.


```{r, echo = FALSE}

weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>%
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) %>%
  select(name, id, everything())

weather_df

```

We’ll start with a basic scatterplot to develop our understanding of ggplot’s data + mappings + geoms approach, and build quickly from there.

## Basic scatterplot

```{r, warning = FALSE}
ggplot(weather_df, aes(x = tmin, y = tmax)) + geom_point()

```


The code below could be used instead to produce the same figure. Using this style can be helpful if you want to do some pre-processing before making your plot but don’t want to save the intermediate data.

drop_na removes na values

```{r}
weather_df %>% 
  
  drop_na() %>%
  
  filter(name == "CentralPark_NY") %>%
  
  ggplot(aes(x = tmin, y = tmax)) + 
  
  geom_point()
```

Notice that we try to use good styling practices here as well – new plot elements are added on new lines, code that’s part of the same sequence is indented, we’re making use of whitespace, etc.

You can also save the output of ggplot() to an object and modify / print it later. This is often helpful, although it’s not my default approach to making plots.

```{r, warning = FALSE}
plot_weather = 
  weather_df %>%
  ggplot(aes(x = tmin, y = tmax))

plot_weather + geom_point()
```

## Advanced scatterplots

The basic scatterplot gave some useful information – the variables are related roughly as we’d expect, and there aren’t any obvious outliers to investigate before moving on. We do, however, have other variables to learn about using additional aesthetic mappings.

Let’s start with name, which I can incorporate using the color aesthetic:

```{r, warning = FALSE}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  
  geom_point(aes(color = name))
```


next I’ll add a smooth curve and make the data points a bit transparent. Alpha makes points more or less transparent, goes from 0-1 where 0 is not transparent, 1 is completely

```{r, message = FALSE, warning = FALSE}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = 0.5) + 
  geom_smooth(se = FALSE)
```

The X and Y mappings apply to the whole graphic, but color is currently geom-specific. Sometimes you want or need to do this, but for now I don’t like it. If I’m honest, I’m also having a hard time seeing everything on one plot, so I’m going to add facet based on name as well.

```{r, message = FALSE, warning = FALSE}
ggplot(weather_df, aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)

```

This splits up the three datsets into their own individual scatterplots


However, the relationship between minimum and maximum temperature is now kinda boring, so I’d prefer something that shows the time of year. Also I want to learn about precipitation, so let’s do that.

```{r, message = FALSE, warning = FALSE}
ggplot(weather_df, aes(x = date, y = tmax, color = name)) + 
  geom_point(aes(size = prcp), alpha = 0.5) + 
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name) 
```

Write a code chain that starts with weather_df; focuses only on Central Park, converts temperatures to Fahrenheit, makes a scatterplot of min vs. max temperature, and overlays a linear regression line (using options in geom_smooth())

```{r, message = FALSE}
weather_df %>%
  
  filter(name == "CentralPark_NY") %>%
  mutate(
    tmin_fahr = tmin * (9/5) + 32,
    tmax_fahr = tmax * (9/5) + 32) %>%
  
  ggplot(aes(x = tmin_fahr, y = tmax_fahr)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm", se = FALSE)
  
  

```

se in geom_smooth displays confidence interval so when false this is hidden, lm plots a linear line over the scatterplot

## Univariate plots

Scatterplots are great, but sometimes you need to carefully understand the distribution of single variables 

```{r, message = FALSE, warning = FALSE}
ggplot(weather_df, aes(x = tmax)) + 
  geom_histogram()
```

Can change the bind width and set fill color using aesthetics
```{r, warning = FALSE}
ggplot(weather_df, aes(x = tmax, fill = name)) + 
  geom_histogram(position = "dodge", binwidth = 2) + 
  facet_grid(. ~ name)
```

position = dodge places the bars for ech group side by side 

Now we use density plots

```{r, warning = FALSE}
ggplot(weather_df, aes(x = tmax, fill = name)) + 
  geom_density(alpha = 0.4, adjust = 0.5, color = "blue")
```

The adjust parameter is similar to binwidth parameter, transparency is set to 0.4 to make sure all densities appear, adding geom_rug() to density plot can be helpful way to show raw data in addition to density

```{r, warning = FALSE }
ggplot(weather_df, aes(x = name, y = tmax)) + 
  geom_boxplot()
```

Now we use violin plots

```{r, warning = FALSE}
ggplot(weather_df, aes(x = name, y = tmax)) + 
  geom_violin(aes(fill = name), alpha = 0.5) + 
  stat_summary(fun = "median", color = "blue")
```

Ridge plots were the trendiest plot of 2017, and were a replacement for both boxplots and violin plots. They’re implemented in the ggridges package, and are nice if you have lots of categories in which the shape of the distribution matters.

```{r, message = FALSE, warning = FALSE}
ggplot(weather_df, aes(x = tmax, y = name)) + 
  geom_density_ridges(scale = 0.85)
```


Make plots that compare precipitation across locations. Try a histogram, a density plot, a boxplot, a violin plot, and a ridgeplot; use aesthetic mappings to make your figure readabable


```{r, warning = FALSE}

ggplot(weather_df, aes(x = prcp, fill = name)) + 
  geom_histogram(position = "dodge", binwidth = 2)

ggplot(weather_df, aes(x = prcp, y = name)) + 
  geom_density(alpha = 0.3)

ggplot(weather_df, aes(x = name, y = prcp)) + 
  geom_boxplot()

ggplot(weather_df, aes(x = name, y = prcp)) + 
  geom_violin(aes(fill = name), alpha = 0.5)
```

## Saving and embedding plots

Don't use export button, instead use ggsave

```{r, warning = FALSE}
weather_plot = ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = 0.5)

ggsave("weather_plot.pdf", weather_plot, width = 8, height = 5)
```

Embedding plots in an R Markdown document can also take a while to get used to, because there are several things to adjust. First is the size of the figure created by R, which is controlled using two of the three chunk options fig.width, fig.height, and fig.asp

I prefer a common width and plots that are a little wider than they are tall, so I set options to fig.width = 6 and fig.asp = .6

Second is the size of the figure inserted into your document, which is controlled using out.width or out.height. I like to have a little padding around the sides of my figures, so I set out.width = "90%". I do all this by including the following in a code snippet at the outset of my R Markdown documents.
 

```{r}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = 0.6,
  out.width = "90%"
)
```

What makes embedding figures difficult at first is that things like the font and point size in the figures generated by R are constant – that is, they don’t scale with the overall size of the figure. As a result, text in a figure with width 12 will look smaller than text in a figure with width 6 after both have been embedded in a document. 